Reagents: expressing and composing fine-grained concurrency

### 2011-11-06 paper/paper.tex
writing

\elide{
High level concepts:
  Data abstraction/internal normalized syntactic representation
  Continuations/Commit
  Reactions
  Retry versus Block
  Offers

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming afficianados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{BacktrackCommand}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
reagent and the reaction built up so far, and either completes the reaction and
returns its result (type \lstinline{B}) or one of the failure singletons
(\lstinline{Block} or \lstinline{Retry}).

@@

\begin{figure}
\begin{lstlisting}
def !(a: A): B = tryReact(a, empty, null) match {
  case (_: BacktrackCommand) => {
    val backoff = new Backoff
    val maySync = this.maySync // cache
    def retryLoop(shouldBlock: Boolean): B = {
      val wait = maySync || shouldBlock
      val waiter = if (wait) new Waiter[B](shouldBlock) 
                   else null

      tryReact(a, empty, waiter) match {
        case (bc: BacktrackCommand) if wait => {
          bc.bottom(waiter, backoff, snoop(a))
          waiter.tryAbort      // rescind waiter,
          waiter.poll match { // but check if completed
            case Some(ans) => ans.asInstanceOf[B] 
            case None       => retryLoop(bc.isBlock)
          }
        }
        case Retry => backoff.once; retryLoop(false)
        case Block => retryLoop(true)
        case ans   => ans.asInstanceOf[B]
      }
    }
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

@@

>>>>

\elide{
High level concepts:
x Reactions
x Data abstraction/internal normalized syntactic representation
x Retry versus Block
  Continuations/Commit
  Offers

@@

In the implementation, \lstinline{Reagent[A,B]} is an abstract class all of
whose subclasses are hidden.  The subclasses roughly correspond to the
combinator functions (which are responsible for instantiating them), and
instances of the subclasses store the arguments given to their corresponding
combinator.  Each subclass provides an implementation of the
\lstinline{tryReact} method, which is an abstract method of
\lstinline{Reagent[A,B]} with the following signature:\footnote{ Functional
  programming afficianados should regard \lstinline{tryReact} as a kind of
  Kleisli arrow for reagents.}
\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any
\end{lstlisting}
The \lstinline{Any} type in Scala lies at the top of the subtyping hierarchy,
akin to \lstinline{Object} in Java.  Here we are using \lstinline{Any} to
represent a union of the type \lstinline{B} with the type
\lstinline{Failure}, where the latter has just two singleton instances,
\lstinline{Block} and \lstinline{Retry}, corresponding to permanent and
transient failures.

In other words, \lstinline{tryReact} takes the input (type \lstinline{A}) to the
reagent and the reaction built up so far, and either completes the reaction,
returning a result (type \listinline{B}), or fails, returning one of the
failure singletons (\lstinline{Block} or \lstinline{Retry}).

@@

\elide{
def !(a: A): B = tryReact(a, empty, null) match {
  case (_: Failure) => {
    val backoff = new Backoff
    def retryLoop(shouldBlock: Boolean): B = {
      val offer = if (maySync || shouldBlock) 
                    new Offer[B]
                  else null
      tryReact(a, empty, offer) match {
        case (f: Failure) => {
          if (f == Block && offer != null) 
            LockSupport.park 
          else backoff.once

          // rescind offer, but check for completion
          if (offer != null && offer.tryAbort)
            offer.answer
          else
            retryLoop(f == Block)
        }
        case ans   => ans.asInstanceOf[B]
      }
    }
    backoff.once
    retryLoop(false)
  }
  case ans => ans.asInstanceOf[B]
}
}

\begin{figure}
\begin{lstlisting}
def !(a: A): B = {
  val backoff = new Backoff
  def retryLoop(shouldBlock: Boolean): B = {
    if (maySync || shouldBlock) {
      val offer = new Offer[B]
      tryReact(a, empty, offer) match {
        case Block => 
          LockSupport.park
          if (offer.rescind) retryLoop(true)
          else offer.answer
        case Retry =>
          backoff.once
          if (offer.rescind) retryLoop(maySync)
          else offer.answer
      }
    } else {
      tryReact(a, empty, null) match {
        case Block => retryLoop(true)
        case Retry => backoff.once; retryLoop(maySync)
        case ans   => ans.asInstanceOf[B]
      }
    }
  retryLoop(false)
}
\end{lstlisting}
\nocaptionrule
\caption{The \lstinline{!} method, defined in \lstinline{Reagent[A,B]}}
\label{fig:reaction}
\end{figure}
%TODO: note tail recursion

@@

\begin{lstlisting}
  def tryReact(a: A, rx: Reaction, offer: Offer[B]): Any = 
    c(a).tryReact((), rx, offer)
\end{lstlisting}
